---
title: "Climate Perturbations"
description: "Applying monthly perturbations to stochastic weather, and validate precipitation changes."
format:
  html:
    toc: true
    toc-location: left
    toc-title: "Contents"
    toc-depth: 2
    toc-expand: 2
    page-layout: full
    number-sections: true
    code-tools: true
    df-print: tibble
    self-contained: true
execute:
  echo: true
  warning: false
  message: false
  cache: false
  freeze: true
  execute-dir: project
vignette: >
  %\VignetteIndexEntry{Climate Perturbations}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

# Purpose

This vignette demonstrates how to apply **climate perturbations** to daily
weather series using `weathergenr`. The workflow modifies precipitation and
temperature in a controlled way to reflect user-defined stress scenarios while
preserving realistic daily variability. This is useful for hydrological stress
testing, climate sensitivity analysis, and impact modeling.

The examples below perturb the **observed** record, which keeps the mechanics of
each argument visible without a generator run in the way.
That is a demonstration, not the intended workflow.
In a stress test you perturb the *stochastic realizations*, so that scenarios
inherit the generator's low-frequency variability rather than the single
historical sequence.
@sec-realizations shows that chain.

::: {.callout-note}
## Perturbation sits outside the generator
`apply_climate_perturbations()` is deliberately not a stage inside
`run_weather_generator()`.
That entry point evaluates its output against the observed record, and a
perturbed series is meant to depart from observations, so the two do not belong
in one pipeline.
Generate first, then perturb.
:::

# Setup

```{r setup}
library(weathergenr)
library(ggplot2)

ncfile <- system.file("extdata", "ntoum_era5_data.nc", package = "weathergenr")
ncdata <- read_netcdf(ncfile)
```

# Define climate scenario

We define a stress-test scenario representing a drier and warmer climate:
30% reduction in mean precipitation and +2 degrees Celsius temperature increase.

```{r climate_factors}
obs_date <- ncdata$date
obs_month <- as.integer(format(obs_date, "%m"))
obs_year <- as.integer(format(obs_date, "%Y"))
obs_year_idx <- obs_year - min(obs_year) + 1L
n_years <- max(obs_year_idx)

# Monthly perturbation vectors (length 12)
precip_mean_change <- rep(0.7, 12)
precip_var_change <- rep(1.0, 12)
temp_mean_change <- rep(2, 12)

# Expand to matrices (n_years x 12) for quantile mapping
precip_mean_change_m <- matrix(precip_mean_change, nrow = n_years, ncol = 12, byrow = TRUE)
precip_var_change_m <- matrix(precip_var_change, nrow = n_years, ncol = 12, byrow = TRUE)
```

# Preview quantile mapping

Before applying perturbations across all grid cells, we preview the
quantile-mapping diagnostics on a single cell to verify target changes are
achieved.

```{r quantile_mapping_preview}
precip_qm <- adjust_precipitation_qm(
  precip = ncdata$data[[1]]$precip,
  mean_factor = precip_mean_change_m,
  var_factor = precip_var_change_m,
  scale_var_with_mean = TRUE,
  exaggerate_extremes = FALSE,
  enforce_target_mean = TRUE,
  month = obs_month,
  year = obs_year_idx,
  min_events = 10,
  validate_output = TRUE,
  diagnostics = TRUE,
  verbose = FALSE
)

print(precip_qm$diagnostics)
```

# Apply climate perturbations

The `apply_climate_perturbations()` function processes all grid cells. It
accepts a list of data frames (`data`), grid metadata (`grid`), and a date
vector (`date`). Monthly perturbation factors control precipitation mean/variance
scaling and temperature shifts. When `diagnostic = TRUE`, the function returns
both the perturbed data and validation diagnostics.

```{r apply_perturbations}
rlz_future <- apply_climate_perturbations(
  data = ncdata$data,
  grid = ncdata$grid,
  date = ncdata$date,
  precip_mean_factor = precip_mean_change,
  precip_var_factor = precip_var_change,
  temp_delta = temp_mean_change,
  precip_occurrence_factor = NULL,
  precip_intensity_threshold = 0,
  # Diurnal temperature range. NULL leaves it untouched, which is what warming
  # alone does: temp_delta shifts temp, temp_min and temp_max together. That
  # matters if you go on to compute PET -- Hargreaves goes as the square root of
  # the range, so warming on its own moves PET by only about 2.3% per degree and
  # the range decides how much more. There is no defensible default; supply
  # values for the region you are studying, or vary it as a stress dimension.
  temp_range_factor = NULL,
  temp_transient = TRUE,
  precip_transient = TRUE,
  precip_occurrence_transient = TRUE,
  compute_pet = FALSE,
  pet_method = "hargreaves",
  qm_fit_method = "mme",
  scale_var_with_mean = TRUE,
  exaggerate_extremes = FALSE,
  extreme_prob_threshold = 0.95,
  extreme_k = 1.2,
  enforce_target_mean = TRUE,
  precip_cap_mm_day = NULL,
  precip_floor_mm_day = NULL,
  precip_cap_quantile = NULL,
  seed = 42,
  diagnostic = TRUE,
  verbose = FALSE
)
```

The output contains `data` (list of perturbed data frames) and `diagnostic`
(list of per-grid validation objects, precipitation only).

# Visualizations

## Monthly precipitation

```{r plot_monthly_precip}
obs_cell <- ncdata$data[[1]]
sim_cell <- rlz_future$data[[1]]

obs_monthly <- tapply(obs_cell$precip, obs_month, mean, na.rm = TRUE)
sim_monthly <- tapply(sim_cell$precip, obs_month, mean, na.rm = TRUE)

plot_data <- data.frame(
  Month = factor(month.abb[1:12], levels = month.abb),
  Observed = as.numeric(obs_monthly),
  Perturbed = as.numeric(sim_monthly)
)
plot_data_long <- tidyr::pivot_longer(
  plot_data, cols = c("Observed", "Perturbed"),
  names_to = "Series", values_to = "Precipitation"
)

ggplot(plot_data_long, aes(x = Month, y = Precipitation, fill = Series)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_fill_manual(values = c(Observed = "steelblue", Perturbed = "coral")) +
  labs(
    title = "Monthly Mean Precipitation",
    subtitle = "Observed vs. perturbed (30% reduction scenario)",
    y = "Precipitation (mm/day)", x = NULL
  ) +
  theme_bw() +
  theme(legend.position = "bottom")
```

## Distribution comparison

```{r plot_distribution}
wet_obs <- obs_cell$precip[obs_cell$precip > 0]
wet_sim <- sim_cell$precip[sim_cell$precip > 0]

dist_data <- data.frame(

  Precipitation = c(wet_obs, wet_sim),
  Series = factor(
    c(rep("Observed", length(wet_obs)), rep("Perturbed", length(wet_sim))),
    levels = c("Observed", "Perturbed")
  )
)

ggplot(dist_data, aes(x = Precipitation, color = Series, fill = Series)) +
  geom_density(alpha = 0.3, linewidth = 1) +
  scale_color_manual(values = c(Observed = "steelblue", Perturbed = "coral")) +
  scale_fill_manual(values = c(Observed = "steelblue", Perturbed = "coral")) +
  scale_x_continuous(limits = c(0, quantile(c(wet_obs, wet_sim), 0.98))) +
  labs(
    title = "Wet-Day Precipitation Distribution",
    x = "Precipitation (mm/day)", y = "Density"
  ) +
  theme_bw() +
  theme(legend.position = "bottom")
```

## Temperature time series

```{r plot_temperature}
plot_idx <- seq_len(nrow(obs_cell))

temp_ts <- data.frame(
  Day = plot_idx,
  Observed = obs_cell$temp[plot_idx],
  Perturbed = sim_cell$temp[plot_idx]
)
temp_ts_long <- tidyr::pivot_longer(
  temp_ts, cols = c("Observed", "Perturbed"),
  names_to = "Series", values_to = "Temperature"
)

ggplot(temp_ts_long, aes(x = Day, y = Temperature, color = Series)) +
  geom_line(alpha = 0.7) +
  scale_color_manual(values = c(Observed = "steelblue", Perturbed = "coral")) +
  labs(
    title = "Daily Temperature Time Series",
    subtitle = "Observed vs. +2 deg C shift scenario",
    x = "Day", y = "Temperature (deg C)"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

# Perturbing generated realizations {#sec-realizations}

Everything above perturbed the observed record.
A stress test perturbs the generator's output instead, so each scenario carries
the interannual variability the generator produced rather than the one sequence
history happened to deliver.

The generator returns *resampled dates*, not values, so there is one join
between the stages.
`prepare_evaluation_data()` performs it and returns exactly the per-cell shape
`apply_climate_perturbations()` expects, so no glue code is needed:

```{r}
#| label: perturb-realizations
#| eval: false
vars <- c("precip", "temp", "temp_min", "temp_max")

gen <- generate_weather(
  obs_data       = ncdata$data,
  obs_grid       = ncdata$grid,
  obs_dates      = ncdata$date,
  vars           = vars,
  n_years        = 20,
  n_realizations = 2,
  warm_var       = "precip",
  out_dir        = tempdir(),
  seed           = 123
)

# Join resampled dates back to daily values
eval_data <- prepare_evaluation_data(
  gen_output = gen,
  obs_data   = ncdata$data,
  obs_dates  = ncdata$date,
  grid_ids   = seq_along(ncdata$data),
  variables  = vars,
  verbose    = FALSE
)

# Perturb one realization
sim <- eval_data$sim_data[[1]]

rlz_future <- apply_climate_perturbations(
  data               = sim,
  grid               = ncdata$grid,
  date               = sim[[1]]$date,
  precip_mean_factor = rep(0.7, 12),
  precip_var_factor  = rep(1.0, 12),
  temp_delta         = rep(2, 12),
  compute_pet        = FALSE,
  seed               = 42,
  diagnostic         = TRUE,
  verbose            = FALSE
)
```

Loop over `eval_data$sim_data` to perturb the whole ensemble, and over a set of
factor combinations to build a stress-test grid.

## Three things that catch people

**`vars` must include `temp_min` and `temp_max`.**
`generate_weather()` requires only `precip` and `temp`, but
`apply_climate_perturbations()` requires all four.
Ask for them at the generator, since `prepare_evaluation_data()` can only
extract variables that came through.

**Take the date vector from `sim`, not from `gen$dates`.**
`prepare_evaluation_data()` drops incomplete years, so its frames can be shorter
than `gen$dates`.
Passing `gen$dates` then fails on a row-count mismatch.
`sim[[1]]$date` is always aligned with the data it came from.

**Matrix factors are sized by calendar years spanned.**
The `n_years x 12` form is indexed by calendar year, not by the generator's
`n_years`.
Under a water year (`year_start_month > 1`) a 20-year series spans 21 calendar
years and needs a `21 x 12` matrix.
Use `length(unique(format(sim[[1]]$date, "%Y")))` to size it, or supply a
length-12 vector, which applies to every year and sidesteps the question.

# Use cases and extensions

## Step changes

For immediate perturbations instead of transient ramps:

```{r step_changes, eval=FALSE}
rlz_step <- apply_climate_perturbations(
  data = ncdata$data, grid = ncdata$grid, date = ncdata$date,
  precip_mean_factor = precip_mean_change,
  precip_var_factor = precip_var_change,
  temp_delta = temp_mean_change,
  temp_transient = FALSE,
  precip_transient = FALSE,
  diagnostic = TRUE
)
```

## Occurrence perturbations

To modify wet-day frequency:

```{r occurrence_example, eval=FALSE}
occurrence_factor <- rep(1.0, 12)
occurrence_factor[6:8] <- 0.8  # 20% fewer wet days in summer

rlz_dry_summer <- apply_climate_perturbations(
  data = ncdata$data, grid = ncdata$grid, date = ncdata$date,
  precip_mean_factor = precip_mean_change,
  precip_var_factor = precip_var_change,
  precip_occurrence_factor = occurrence_factor,
  temp_delta = temp_mean_change,
  diagnostic = TRUE
)
```

## PET recalculation

To propagate temperature changes to PET:

```{r pet_example, eval=FALSE}
rlz_with_pet <- apply_climate_perturbations(
  data = ncdata$data, grid = ncdata$grid, date = ncdata$date,
  precip_mean_factor = precip_mean_change,
  precip_var_factor = precip_var_change,
  temp_delta = temp_mean_change,
  compute_pet = TRUE,
  pet_method = "hargreaves",
  diagnostic = TRUE
)
```

## Extreme amplification

To amplify upper-tail changes:

```{r extreme_example, eval=FALSE}
rlz_extreme <- apply_climate_perturbations(
  data = ncdata$data, grid = ncdata$grid, date = ncdata$date,
  precip_mean_factor = precip_mean_change,
  precip_var_factor = precip_var_change,
  temp_delta = temp_mean_change,
  exaggerate_extremes = TRUE,
  extreme_prob_threshold = 0.95,
  extreme_k = 1.2,
  diagnostic = TRUE
)
```

# Session info

```{r session_info}
sessionInfo()
```
