---
title: "Precipitation Perturbation Parameters"
description: "How the precipitation perturbation arguments change the outcome, measured against a common baseline on one stylized site."
format:
  html:
    toc: true
    toc-location: left
    toc-title: "Contents"
    toc-depth: 3
    toc-expand: 2
    page-layout: full
    number-sections: true
    code-tools: true
    df-print: tibble
    self-contained: true
execute:
  echo: true
  warning: false
  message: false
  cache: false
  freeze: true
  execute-dir: project
vignette: >
  %\VignetteIndexEntry{Precipitation Perturbation Parameters}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

# Purpose

[Climate Perturbations](climate_perturbation.html) shows *how* to apply a
perturbation.
This article asks a different question: given a target change, what do the
remaining arguments actually do to the result?

Every scenario below imposes the same headline change — a 20% reduction in
wet-day precipitation — and varies one argument at a time.
Holding the target fixed is the point: differences between panels are the
argument's doing, not the scenario's.

The site is stylized rather than real, so the mechanisms are visible without a
region's idiosyncrasies on top.
Everything runs in a few seconds and no weather-generator call is involved.

::: {.callout-note}
## What is not covered
`compute_pet` and `pet_method` are driven by temperature, not precipitation, and
are covered in [Climate Perturbations](climate_perturbation.html).
`precip_cap_mm_day`, `precip_floor_mm_day` and `precip_cap_quantile` are safety
rails applied after everything here, and do nothing until a bound is hit.
:::

# Setup

```{r}
#| label: setup
library(weathergenr)
library(ggplot2)
library(dplyr)
library(tidyr)

theme_set(theme_weathergenr())
```

## A stylized site

Thirty years of daily precipitation on a 365-day calendar, with a wet season
from November to March, gamma-distributed wet-day amounts, and a slow
interannual oscillation.
That last ingredient matters: without year-to-year variability, one of the
arguments below has nothing to act on and looks inert when it is not.

```{r}
#| label: stylized-site
n_years <- 30L
start_year <- 2000L

date <- as.Date("2000-01-01") + 0:(365L * n_years - 1L)
month <- as.integer(format(date, "%m"))
year_idx <- as.integer(format(date, "%Y")) - start_year + 1L

wet_prob <- c(0.55, 0.50, 0.45, 0.25, 0.15, 0.10,
              0.08, 0.10, 0.18, 0.30, 0.45, 0.55)[month]
wet_scale <- c(9, 8.5, 7.5, 5, 3.5, 2.5,
               2, 2.5, 4, 6, 8, 9)[month]

# A slow wet/dry oscillation, so the series has interannual structure.
ann_mult <- 1 + 0.35 * sin(2 * pi * seq_len(n_years) / 11)

set.seed(20260817)
precip <- numeric(length(date))
is_wet <- runif(length(date)) < wet_prob * ann_mult[year_idx]
precip[is_wet] <- rgamma(
  sum(is_wet),
  shape = 0.75,
  scale = wet_scale[is_wet] * ann_mult[year_idx][is_wet]
)
```

```{r}
#| label: site-summary
tibble(
  days           = length(precip),
  wet_days       = sum(precip > 0),
  wet_fraction   = mean(precip > 0),
  annual_mm      = mean(tapply(precip, year_idx, sum)),
  wetday_mean_mm = mean(precip[precip > 0])
)
```

## Instruments and metrics

The shape arguments belong to `adjust_precipitation_qm()`, the quantile mapper
that `apply_climate_perturbations()` calls internally.
Going straight to it keeps the grid and temperature scaffolding out of the way.
The two `*_transient` arguments exist only on the wrapper, so
@sec-transient switches instruments.

Note the one interface difference: the wrapper accepts a length-12 vector and
expands it, while the mapper requires an `n_years x 12` matrix.

```{r}
#| label: helpers
# The mapper wants a matrix; expand a monthly vector to one.
as_factor_matrix <- function(v) {
  matrix(v, nrow = n_years, ncol = 12, byrow = TRUE)
}

perturb <- function(mean_factor = 0.8, var_factor = 1, ...) {
  adjust_precipitation_qm(
    precip      = precip,
    month       = month,
    year        = year_idx,
    mean_factor = as_factor_matrix(rep(mean_factor, 12)),
    var_factor  = as_factor_matrix(rep(var_factor, 12)),
    validate_output = FALSE,
    diagnostics     = FALSE,
    verbose         = FALSE,
    ...
  )
}
```

Four metrics, chosen so the arguments separate rather than overlap.
The last two are the ones that matter most, and pooling them — as a single
overall coefficient of variation would — hides exactly the distinction
@sec-two-knobs draws.

```{r}
#| label: metrics
wet_of <- function(x) x[x > 0]

# Groups are (month, year), the unit the perturbation itself works on.
by_group <- function(x) {
  w <- x > 0
  g <- split(x[w], list(month[w], year_idx[w]), drop = TRUE)
  g[lengths(g) >= 5]
}

summarise_series <- function(x, label) {
  g <- by_group(x)
  group_means <- vapply(g, mean, numeric(1))
  tibble(
    scenario    = label,
    wetday_mean = mean(wet_of(x)),
    p99         = unname(quantile(wet_of(x), 0.99)),
    # Spread of daily values inside a (month, year) group, averaged.
    within_cv   = mean(vapply(g, function(v) sd(v) / mean(v), numeric(1))),
    # Spread of the group means themselves: the interannual signal.
    between_cv  = sd(group_means) / mean(group_means)
  )
}
```

The baseline every scenario is measured against applies the 20% reduction with
every other argument at its default.

```{r}
#| label: baseline
baseline <- perturb()

bind_rows(
  summarise_series(precip, "observed"),
  summarise_series(baseline, "baseline (-20%, defaults)")
)
```

# The headline: two ways to make extremes worse {#sec-extremes}

`exaggerate_extremes` amplifies the upper tail.
`enforce_target_mean` rescales each (month, year) group so its wet-day mean hits
the target exactly.
They are usually described separately, but the second decides what the first
*means*.

```{r}
#| label: extremes-grid
grid_2x2 <- expand.grid(
  exaggerate = c(FALSE, TRUE),
  enforce    = c(TRUE, FALSE)
)

do.call(bind_rows, lapply(seq_len(nrow(grid_2x2)), function(i) {
  ex <- grid_2x2$exaggerate[i]
  en <- grid_2x2$enforce[i]
  summarise_series(
    perturb(exaggerate_extremes = ex, enforce_target_mean = en, extreme_k = 1.5),
    sprintf("exaggerate=%s, enforce=%s", ex, en)
  )
}))
```

The target wet-day mean is `r round(0.8 * mean(precip[precip > 0]), 3)` mm.
Three of the four rows sit exactly on it.
The fourth — amplification with `enforce_target_mean = FALSE` — overshoots by
about 13%.

That is the whole interaction.
Amplifying the tail moves probability mass upward, which *adds* water; the
scenario then delivers a larger mean change than the one requested.
`enforce_target_mean = TRUE` rescales the group back onto its target, so the
same amplification instead *redistributes* water: more in the tail, less in the
body, at the mean you asked for.

::: {.callout-important}
## Which one you want is a modelling choice
Redistribution at a fixed mean is the usual intent for a stress test, because it
keeps the mean-change axis interpretable.
If you leave `enforce_target_mean = FALSE` while amplifying, the realized mean
change is no longer the number in your scenario table.
:::

```{r}
#| label: fig-extremes
#| fig-cap: "Wet-day exceedance curves. With the mean enforced, tail amplification pivots the distribution about its mean; without it, the whole curve shifts up."
#| fig-alt: "Log-scaled exceedance probability curves for four combinations of exaggerate_extremes and enforce_target_mean, showing the unenforced amplified case lying above the others across the full range."
#| fig-width: 8
#| fig-height: 4.5
exceedance <- function(x, label) {
  v <- sort(wet_of(x), decreasing = TRUE)
  tibble(value = v, prob = seq_along(v) / (length(v) + 1), scenario = label)
}

bind_rows(
  exceedance(baseline, "no amplification"),
  exceedance(perturb(exaggerate_extremes = TRUE, enforce_target_mean = TRUE,
                     extreme_k = 1.5), "amplified, mean enforced"),
  exceedance(perturb(exaggerate_extremes = TRUE, enforce_target_mean = FALSE,
                     extreme_k = 1.5), "amplified, mean free")
) |>
  filter(prob < 0.3) |>
  ggplot(aes(prob, value, colour = scenario)) +
  geom_line(linewidth = 0.9) +
  scale_x_log10(labels = scales::percent) +
  scale_colour_viridis_d(end = 0.85) +
  labs(
    title = "Enforcing the target mean turns amplification into redistribution",
    x = "Exceedance probability", y = "Wet-day precipitation (mm)",
    colour = NULL
  ) +
  theme(legend.position = "bottom")
```

## How hard to amplify: `extreme_k` {#sec-extreme-k}

With the mean enforced, `extreme_k` is a clean dial.
Values above 1 amplify, values below 1 dampen, and 1 is the identity.

```{r}
#| label: k-sweep
k_values <- c(0.6, 0.8, 1.0, 1.2, 1.5, 2.0, 2.5)

k_sweep <- do.call(bind_rows, lapply(k_values, function(k) {
  summarise_series(
    perturb(exaggerate_extremes = TRUE, enforce_target_mean = TRUE, extreme_k = k),
    sprintf("k = %.1f", k)
  )
})) |>
  mutate(k = k_values, .before = 1)

k_sweep
```

The wet-day mean does not move across the whole sweep, while the 99th percentile
roughly doubles.
That is the response surface a stress test wants from a tail dimension: one
axis, no side effects on the others.

```{r}
#| label: fig-k-sweep
#| fig-cap: "Tail severity against extreme_k, with the wet-day mean pinned by enforce_target_mean."
#| fig-alt: "Two-panel plot: the 99th percentile rises steadily with extreme_k while the wet-day mean stays flat."
#| fig-width: 8
#| fig-height: 4
k_sweep |>
  select(k, `99th percentile` = p99, `wet-day mean` = wetday_mean) |>
  pivot_longer(-k) |>
  ggplot(aes(k, value)) +
  geom_line(linewidth = 0.9, colour = "steelblue") +
  geom_point(size = 2, colour = "steelblue") +
  facet_wrap(~name, scales = "free_y") +
  expand_limits(y = 0) +
  labs(
    title = "extreme_k moves the tail and leaves the mean alone",
    x = "extreme_k", y = "mm/day"
  )
```

## Where the tail starts: `extreme_prob_threshold`

`extreme_prob_threshold` sets the probability above which amplification applies.
Its effect is less intuitive than `extreme_k`, and worth seeing before use.

```{r}
#| label: u0-sweep
u0_values <- c(0.80, 0.90, 0.95, 0.99)

tail_quantiles <- function(x, label) {
  tibble(
    scenario = label,
    p90  = unname(quantile(wet_of(x), 0.90)),
    p99  = unname(quantile(wet_of(x), 0.99)),
    p999 = unname(quantile(wet_of(x), 0.999)),
    wetday_mean = mean(wet_of(x))
  )
}

bind_rows(
  tail_quantiles(baseline, "no amplification"),
  do.call(bind_rows, lapply(u0_values, function(u0) {
    tail_quantiles(
      perturb(exaggerate_extremes = TRUE, enforce_target_mean = TRUE,
              extreme_k = 1.5, extreme_prob_threshold = u0),
      sprintf("u0 = %.2f", u0)
    )
  }))
)
```

Read the columns, not the rows.
Lowering the threshold widens the affected region: at `u0 = 0.80` even the 90th
percentile rises, because amplification has reached into the body of the
distribution.
Raising it concentrates the effect — at `u0 = 0.99` the 90th and 99th
percentiles are barely touched while the 99.9th still moves.

The 99th percentile is not monotone across this sweep, which looks odd until you
notice it sits *at* the threshold for the last row.
With the mean enforced, widening the amplified region also forces a larger
compensating rescale, so a given percentile can fall even as the far tail rises.
Pick the threshold to match the return period you care about and read the
quantile that matches it, rather than treating `u0` as a severity dial —
that is `extreme_k`'s job.

# Two knobs, two kinds of variability {#sec-two-knobs}

`scale_var_with_mean` and `enforce_target_mean` both sound like they control
"variability", and a single pooled coefficient of variation would make them look
like rivals.
They are not: they act on different things, and the two `*_cv` columns separate
them cleanly.

```{r}
#| label: two-knobs
knob_grid <- expand.grid(enforce = c(TRUE, FALSE), scale_var = c(TRUE, FALSE))

bind_rows(
  summarise_series(precip, "observed"),
  do.call(bind_rows, lapply(seq_len(nrow(knob_grid)), function(i) {
    en <- knob_grid$enforce[i]
    sv <- knob_grid$scale_var[i]
    summarise_series(
      perturb(enforce_target_mean = en, scale_var_with_mean = sv),
      sprintf("enforce=%s, scale_var=%s", en, sv)
    )
  }))
)
```

`within_cv` responds only to `scale_var_with_mean`, and `between_cv` only to
`enforce_target_mean`.

**`scale_var_with_mean` preserves the shape of a wet day.**
When `TRUE`, the effective variance factor becomes
`var_factor * mean_factor^2`, so the standard deviation scales with the mean and
the coefficient of variation within a group is unchanged — it matches the
observed value exactly.
When `FALSE`, a `var_factor` of 1 holds absolute variance while the mean drops
20%, which inflates relative variability.
Whether that is what you want depends on whether your scenario is "the same
weather, less of it" or "the same absolute spread around a lower mean".

**`enforce_target_mean` costs interannual variability.**
Pinning each group to `climatological mean x factor` means every year receives
the same target for a given month, so year-to-year differences in wet-day mean
are largely erased: `between_cv` falls from the observed value to roughly half
of it.

::: {.callout-warning}
## The default trades away the variability the generator produces
`enforce_target_mean = TRUE` is the default, and it is the right default for
hitting a scenario target exactly.
But interannual variability is precisely what the WARM component of this package
exists to generate, and pinning every year to a climatological target discards a
large part of it.

If your application is sensitive to sequences — multi-year drought, storage
carry-over, anything with memory — check `between_cv` on your own output before
accepting the default, and consider `enforce_target_mean = FALSE` with the
realized mean change verified after the fact rather than imposed.
:::

```{r}
#| label: fig-interannual
#| fig-cap: "Annual mean wet-day precipitation. Enforcing the target mean flattens the interannual signal the observed series carries."
#| fig-alt: "Time series of annual mean wet-day precipitation over thirty years, showing the observed oscillation preserved when enforce_target_mean is FALSE and largely flattened when it is TRUE."
#| fig-width: 8
#| fig-height: 4.5
annual_wetmean <- function(x, label) {
  tibble(
    year = start_year + seq_len(n_years) - 1L,
    value = vapply(split(x, year_idx), function(v) mean(v[v > 0]), numeric(1)),
    scenario = label
  )
}

bind_rows(
  annual_wetmean(precip, "observed"),
  annual_wetmean(perturb(enforce_target_mean = TRUE), "enforce_target_mean = TRUE"),
  annual_wetmean(perturb(enforce_target_mean = FALSE), "enforce_target_mean = FALSE")
) |>
  ggplot(aes(year, value, colour = scenario)) +
  geom_line(linewidth = 0.9) +
  scale_colour_viridis_d(end = 0.85) +
  labs(
    title = "Enforcing the target mean flattens year-to-year variability",
    x = NULL, y = "Annual mean wet-day precipitation (mm)", colour = NULL
  ) +
  theme(legend.position = "bottom")
```

# Fitting: `qm_fit_method`

`qm_fit_method` chooses how the monthly Gamma distributions are estimated —
`"mme"` (moment matching, the default) or `"mle"`.

```{r}
#| label: fit-method
fit_mme <- perturb(fit_method = "mme")
fit_mle <- perturb(fit_method = "mle")

bind_rows(
  summarise_series(fit_mme, "mme"),
  summarise_series(fit_mle, "mle")
)
```

The two rows are identical at the printed precision, so it is worth looking at
the difference directly rather than at a rounded table:

```{r}
#| label: fit-method-difference
tibble(
  wetday_mean_difference = mean(wet_of(fit_mle)) - mean(wet_of(fit_mme)),
  largest_daily_difference = max(abs(fit_mle - fit_mme)),
  as_fraction_of_wetday_mean =
    max(abs(fit_mle - fit_mme)) / mean(wet_of(fit_mme))
)
```

The wet-day means differ in the fourth decimal, and the largest single-day
difference is a small fraction of a typical wet day.
That is the honest result: on a well-behaved Gamma sample the two estimators
effectively coincide.
This is a robustness setting, not a scenario dimension — it earns attention when
fitting is hard (short records, months with few wet days, heavily skewed
samples), not as something to vary across a stress-test ensemble.
`min_events`, which controls how many wet days a month needs before it is fitted
at all, matters more often.

# Transient versus step change {#sec-transient}

The `*_transient` arguments live on `apply_climate_perturbations()`, so this
section uses the wrapper.
That needs grid metadata and temperature columns, which the mapper did not.

```{r}
#| label: wrapper-setup
grid <- data.frame(id = 1L, lat = 0)

set.seed(7)
cell <- data.frame(
  precip   = precip,
  temp     = rnorm(length(date), 25, 3),
  temp_min = rnorm(length(date), 20, 3)
)
cell$temp_max <- pmax(cell$temp_min + 5, rnorm(length(date), 30, 3))
```

A transient factor is specified by its **end state**, not by a per-year path.
`transient = TRUE` ramps linearly from 1 to `2f - 1`, so the ramp averages to
`f` over the period.
A request for `f = 0.7` therefore means "ending 40% down, averaging 30% down",
not "30% down every year".

Occurrence shows this most clearly, because wet-day frequency is not touched by
the mean-enforcement machinery above.

```{r}
#| label: transient-occurrence
wet_fraction_by_year <- function(x) {
  vapply(split(x, year_idx), function(v) mean(v > 0), numeric(1))
}
observed_fraction <- wet_fraction_by_year(precip)

run_occurrence <- function(transient) {
  out <- apply_climate_perturbations(
    data = list(cell), grid = grid, date = date,
    precip_mean_factor       = rep(1, 12),
    precip_occurrence_factor = rep(0.7, 12),
    temp_delta               = rep(0, 12),
    precip_transient             = FALSE,
    precip_occurrence_transient  = transient,
    temp_transient               = FALSE,
    compute_pet = FALSE, diagnostic = FALSE, verbose = FALSE, seed = 1L
  )
  tibble(
    year = start_year + seq_len(n_years) - 1L,
    ratio = as.numeric(wet_fraction_by_year(out[[1]]$precip) / observed_fraction),
    scenario = if (transient) "transient (ramp)" else "step change"
  )
}

occurrence_paths <- bind_rows(run_occurrence(FALSE), run_occurrence(TRUE))

occurrence_paths |>
  group_by(scenario) |>
  summarise(
    first_year = first(ratio), last_year = last(ratio), period_mean = mean(ratio)
  )
```

Both average to the requested 0.7.
The step change holds it flat; the ramp starts at no change and ends near 0.4.

```{r}
#| label: fig-transient
#| fig-cap: "Wet-day frequency relative to observed, for a step change and a transient ramp requesting the same 0.7 factor."
#| fig-alt: "Two time series of the wet-day frequency ratio over thirty years: a flat line near 0.7 for the step change, and a declining line from 1.0 to about 0.4 for the transient ramp."
#| fig-width: 8
#| fig-height: 4
ggplot(occurrence_paths, aes(year, ratio, colour = scenario)) +
  geom_line(linewidth = 0.9) +
  geom_hline(yintercept = 0.7, linetype = "dashed", colour = "grey40") +
  scale_colour_viridis_d(end = 0.7) +
  expand_limits(y = 0) +
  labs(
    title = "Same requested factor, two different paths",
    subtitle = "Dashed line marks the requested 0.7",
    x = NULL, y = "Wet-day frequency / observed", colour = NULL
  ) +
  theme(legend.position = "bottom")
```

## A transient factor takes a vector, not a year-varying matrix

Because the ramp is built from the end state, only the first row of a factor is
read in transient mode.
Supplying a year-varying `n_years x 12` matrix *and* setting the transient flag
is therefore a contradiction, and is an error rather than a silent truncation:

```{r}
#| label: transient-guard
#| error: true
varying <- matrix(0.5, nrow = n_years, ncol = 12)
varying[1, ] <- 1.0

apply_climate_perturbations(
  data = list(cell), grid = grid, date = date,
  precip_mean_factor = varying,
  temp_delta = rep(0, 12),
  precip_transient = TRUE,
  compute_pet = FALSE, diagnostic = FALSE, verbose = FALSE
)
```

Pass a length-12 vector and let the ramp build the path, or set the transient
flag to `FALSE` to have the matrix applied year by year.
A matrix whose rows are all identical is equivalent to a vector, so code that
expanded a vector to a constant matrix is unaffected.

## Intensity transients interact with the enforced mean

The same ramp applied to intensity is harder to read, and the reason is
@sec-two-knobs.

```{r}
#| label: transient-intensity
observed_wetmean <- vapply(split(precip, year_idx),
                           function(v) mean(v[v > 0]), numeric(1))

run_intensity <- function(transient, enforce) {
  out <- apply_climate_perturbations(
    data = list(cell), grid = grid, date = date,
    precip_mean_factor  = rep(0.7, 12),
    temp_delta          = rep(0, 12),
    precip_transient    = transient,
    temp_transient      = FALSE,
    enforce_target_mean = enforce,
    compute_pet = FALSE, diagnostic = FALSE, verbose = FALSE, seed = 1L
  )
  realised <- vapply(split(out[[1]]$precip, year_idx),
                     function(v) mean(v[v > 0]), numeric(1))
  tibble(
    scenario = sprintf("transient=%s, enforce=%s", transient, enforce),
    year_1   = unname(realised / observed_wetmean)[1],
    year_30  = unname(realised / observed_wetmean)[n_years],
    period_mean = mean(realised / observed_wetmean)
  )
}

bind_rows(
  run_intensity(FALSE, TRUE), run_intensity(TRUE, TRUE),
  run_intensity(FALSE, FALSE), run_intensity(TRUE, FALSE)
)
```

With `enforce_target_mean = FALSE` the step change sits at exactly 0.70 in every
year, and the ramp tracks the intended path.
With it `TRUE`, neither ratio matches: each group is pinned to
`climatological mean x factor`, so the ratio against *that year's* observed mean
is above the factor in dry years and below it in wet ones.

Neither is wrong — they are different definitions of the target.
`enforce_target_mean = TRUE` means "the wet-day mean shall be 70% of
climatology"; `FALSE` means "each year gets 70% of whatever it had".
For a transient scenario, where the point is a path through time, the second is
usually the intended reading.

# Choosing settings

A summary of what the article establishes, rather than new results.

| Argument | Controls | Reach for it when |
|---|---|---|
| `exaggerate_extremes`, `extreme_k` | Tail severity at fixed mean | Extremes are a stress dimension of their own |
| `extreme_prob_threshold` | How much of the distribution counts as tail | The return period of interest is known |
| `enforce_target_mean` | Whether the mean is exact or emergent | `TRUE` to hit a target; `FALSE` to keep interannual structure |
| `scale_var_with_mean` | Wet-day shape under a mean change | `TRUE` for "same weather, less of it" |
| `qm_fit_method` | Gamma estimation | Fitting is hard; not a scenario axis |
| `precip_transient`, `precip_occurrence_transient` | Step change versus ramp to an end state | The scenario is a path, not a level |

Two defaults are worth a deliberate decision rather than acceptance.
`enforce_target_mean = TRUE` buys exact targeting at the cost of interannual
variability, which is the thing the generator upstream exists to produce.
And `scale_var_with_mean = TRUE` is a statement that a drier climate has
proportionally similar wet days — defensible, and still a statement.

# Session info

```{r}
#| label: session-info
#| echo: false
sessionInfo()
```
